---
title: "Inferential Statistics on US General Social Survey (GSS) dataset"
author: "Alessandro Pelliccioni"
date: June 5, 2024
---

## 1. Introduction

This project was developed during the study of the Coursera course "Inferential 
Statistics" by Duke University. The dataset in use id the US General Social Survey
cumulative file, which includes years 1972-2012. This projects aims to answer the
following questions using statistical inference tests:

1. Is there a relationship between political view (think of self as liberal or
   conservative) and political party affiliation?

2. Differences in unemployement rate between genders during the years

3. Race representation among unemployemed

4. Income disparity baser on region of interview

## 2. Dataset Description

GSS dataset is a Coursera extract of the original material published on the 
website: https://www.norc.org/research/projects/gss.html.

Description (source GSS project's webssite): In 1972, NORC received a grant from
the National Science Foundation (NSF) to build equity and accessibility within 
social science research. James A. Davis at the University of Chicago and thought
leaders at NORC aimed to create a high-quality, unbiased, and easily accessible
public opinion survey that cataloged America’s thoughts, feelings, and opinions 
over time. That instrument, the General Social Survey, is now among NORC’s 
signature projects. Over five decades, the GSS has compiled data from thousands 
of Americans about their evolving characteristics and attitudes.

This modified version of the GSS dataset consists of 57061 observations with 
114 variables, collected from year 1972 and 2012. Each variable represents the 
answer to a specific question asked to the interviewed sample. 
Not every question is applicable to every respondent.

The survey is conducted by random sampling, the results from this project 
can be generalized to the entire US population. The statistical tests performed
in this project can provide only a hint to a correlation between the variables, 
non a proof of casual relationship.

### 2.1 Load packages and data

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(lattice)
```

```{r load-data}
load("gss.Rdata")
```

## 3. Is there a relationshio between political view and political party 
	affiliation?

Variables analyzed to answer this question:

1. **polviews**: A categorical variable, a seven-point scale, on which the 
	political views that people might hold are arranged from extremely 
	liberal - point 1 - to extremely conservative - point 7.

2. **partyid**: A categorical variable, an eight-point scale, on which the
	political affiliation can be classified as either: 0 - strong Democrat
	1 - not strong Democrat, 2 - Independent near Democrat, 3 - Independent
	4 - Independent near Republican, 5 - not strong Republican, 
	6 - strong Republican, 7 - Other party.

In this study I will focus only in latest trends, so I will extract records
from years 2002-2012 and store them in a new dataset. I will also remove rows
with missing values in either variables.

```{r}
gss %>%
  filter(year > 2002 & 
           !is.na(polviews) &
           !is.na(partyid)) %>%
  select(polviews,partyid)  -> gss_pv

dim(gss_pv)
```

The new dataset has a total 11370 rows from the starting 57061.

### 3.1 Exploratory data analysis

```{r}
ggplot(data = gss_pv, aes(x = gss_pv$polviews))+ geom_bar() +
labs(x = "Political view") +
theme(axis.text.x = element_text(angle=60, hjust = 1))
```

We can see from this plot that the political views are normally distribuited
around the central "moderate" value. (see file Political_View_Barplot.pdf)

```{r}
ggplot(data = gss_pv, aes(x = gss_pv$partyid))+ geom_bar() +
labs(x = "Political affiliation") +
theme(axis.text.x = element_text(angle=60, hjust = 1)) 
```

we can see from this plot that the political affiliation variable is more
uniformly distribuited, except with a minimun count for "Other", and 
maximums on the "Strong Democrat", "Not strong Democrat" and "Independent"
counts (see file Politica_affiliation_variable_barplot.pdf)

```{r}
g <- ggplot(data = gss_pv, aes(x = gss_pv$polviews))
g <- g + geom_bar(aes(fill = partyid), position = "dodge")
g + theme(axis.text.x = element_text(angle=60,hjust = 1))
```

Putting the two variables together we see that the moderate group contains most 
values and is quite equalli represented by political affiliation, except for 
sharp peaks of "Not strongly Democrats" and "Independent". The "Liberal" tail
seems to be dominated by democrats, and the "Conservative" tail by republicats,
(see file Political_affiliation_and_views.pdf).

Mosaic plots provide a good illustration of the relationship between two 
categorical variables. The area of the tiles is proportional to the value 
counts within a group. When tiles across groups all have same areas, 
it indicates independence between the variables.

```{r}
plot(table(gss_pv), main = "Political affiliation VS political views",
	xlab = "Political views", ylab = "Political affiliation")
```

The mosaic plot suggests an association of extreme liberal ideas with strong 
democrat party affiliation and extreme conservative ideas with strong 
republican assosiation. With the majority of interviewed being in the moderate
political view category, which is quite evenly distribuited among
political parties affiliation. (see file Mosaic_plot.pdf)

### 3.2 Inference

Hypothesis

**Null hypothesis:** The frequency of political party affiliation and 
the frequency of political views are independent.

**Alternative hypothesis:** The frequency of political party affiliation and 
the frequency of political views are independent.

Check Conditions for a valid statistical analysis

1. **Independence**: the GSS dataset is collected from a random sample survey. 
We safely assume that the interviews are independent.

2. **Sample Size**: samples are obtained without replacement, and must be
less than 10% of the population. The dataset in use for this analysis consists
of 11370 entries which is less than 10% of the US population.

3. **Degrees of Freedom**: We have 7 frequency levels for the political views 
variable and 8 frequency levels for the political party affiliation,
As we have two categorical variables each with over 2 levels, we will utilize 
the chi-squared test of independence to test the hypothesis.

4. **Expected Counts**: To perform a valid chi-square test the expected counts
for each cell should be at least 5. This must be still checked:

```{r}
chisq.test(gss_pv$polviews,gss_pv$partyid)$expected
```

From the above table we can see that every cell has an expected count of more
than 5 values. We are allowed to go ahead and perform the chi-squared test.

### 3.3 Chi-Square test of independence

```{r}
chisq.test(gss_pv$polviews,gss_pv$partyid)
```

The chi-squared value is 4130.3 and the corrisponding p-value for 42 degrees
of freedom is much lower than the significance level of 0.05.
We can now remove the dataset to free memory.

```{r}
rm(gss_pv)
```

### 3.4 Findings

We have convincing evidence to reject the null hypothesis in favor 
of the alternative hypothesis that the frequency of values in the different
categories of political views and party affiliation are dependent. 
The study is observational, so we can only establish association 
but not causal links between these two variables.

* * *

## 4. Differences in unemployement rate between genders during the years

Variables analyzed to answer this question:

1. **unemp**: A categorical variable, yes or no scale, answering to the question
	"have you been unemployed for more than a month in the last ten years"

2. **sex**: A categorical variable indicating the gender of the respondent,
	1 - male, 2 - female.

3. **year**: we will check if has there been a change in the proportions during 
	the years in moving from 1972 to 2012. 

Prepare a simpler dataset

```{r}
gss %>%
  filter(!is.na(sex) & !is.na(unemp) & !is.na(year)) %>%
  select(sex,unemp,year) -> gss_job

dim(gss_job)
```

The dataset has been reduced from 57061 to 35507 rows.

### 4.1 Exploratory Data Analysys

```{r}
gss_job %>%
  group_by(sex,year) %>%
  summarise(prop = sum(unemp == "Yes")/n()) -> gss_job_prop

ggplot(data=gss_job_prop, aes(x = year,y = prop)) + geom_smooth(aes(fill=sex))
```

The above figure display the loess regression curves for the unemployed 
proportion in men and women, during the years 1972-2012,
(see file Unemp_curve.pdf).

*Observations*

1. The proportion of unemployed females shows a steadily growin line from around
25% in 1972 to around 35% in 2012.

2. The curve for the males starts higher than the one for females in 1972, grows 
up to a local maximum in 1985, decreases to a local minimum in 2000 and grows
steadily to a maximum above 37,5% in 2012

3. The curve describing male's unemployement rate seems to be all the time above
the female's unemployement rate, hinting to a difference between the genders

### 4.2 Inference

Hypothesis

**Null hypothesis**: The mean difference in proportions of men 
and women who have been unemployed for more than a month in the last ten years,
is zero.

**Alternate hypothesis**: The mean difference in proportions of men 
and women who have been unemployed for more than a month in the last ten years
is greater than zero, with the proportion of unemployed men being higher
than that of unemployed women.

To test the hypothesis, we perform the test for two independent sample 
proportions. 

Check Conditions

1. **Independence**: the GSS dataset is collected from a random sample survey. 
We safely assume that the interviews are independent.

2. **Sample size**: to be able to perform t-test to proportions in two independent
groups, we have to check that the minimal condition of at least 10 entries is met
in each cell of the 2x2 matrix describing gender(f/m) vs unemployed(y/n).
In order to check that, we need to make some calculations:

```{r}
gss_job %>%
  summarise(p_pool = sum(gss_job$unemp == "Yes")/n(), 
            n_1 = sum(sex == "Female"),
            n_2 = sum(sex == "Male"),
            n_1_success = p_pool*n_1,
            n_1_fails = (1-p_pool)*n_1,
            n_2_success = p_pool*n_2,
            n_2_fails = (1-p_pool)*n_2)
```

From the data printed by last function call, the values that fill our 2x2 matrix
are in the form n_1_success meaning proportion of unemployed males and so on,
which are all well above the 10 limit.

### 4.3 Inference for proportion between two independent groups with t-test

We use the function inference in stats package which was in use during the course

```{r}
inference(y = unemp, x = sex, data = gss_job, statistic = "proportion", 
	type = "ht", null = 0, success="Yes", alternative ="greater",
	 method = "theoretical")
```

The result is z = 8.1099 with a p-value of  < 0.0001, which is well the 
significance level of 0.05. This gives us convincing evidence that there is
indeed a difference in the proportion of unemployment between genders,
we can reject the null hypothesis.
We can now remove the dataset to free memory.

```{r}
rm(gss_job)
```

### 4.4 Findings

We have convincing evidence to reject the null hypothesis in favor 
of the alternative hypothesis that the proportion of unemployement is different
between genders, during the years between 1972-2012.
We have a strong evidence that the proportion of unemployed males has been
higher during the years of the study.

* * *

## 5. Race representation among unemployed

Variables analyzed to answer this question:

1. **race**: Categorical variable indicating 1- white, 2 - black, 3 - other.
	
2. **unemp**: A categorical variable, yes or no scale, answering to the question
	"have you been unemployed for more than a month in the last ten years"

Let's prepare a simpler dataset with only these two variables:

```{r}
gss %>%
  filter(!is.na(race) & !is.na(unemp)) %>%
  select(race, unemp) -> gss_re

dim(gss_re)
```

The dataset is reduced to 35507 rows and 2 columns.

### 5.1 Exploratory Data Analysys

Let's calculate what are the proportions of each etnic group in the dataset

```{r}
n_tot <- nrow(gss_re) ## total number of entries

gss_re %>%
  group_by(race) %>%
  summarise(p = n()/n_tot) -> etnic_groups

etnic_groups
```

results are:

1 White         0.832 
2 Black         0.123 
3 Other         0.0448

the "White" group has more of 80% of the share of the dataset.
Let's calculate the expected proportion if there wouldn't be difference in
unemployment rate between the different etnic groups.

```{r}
n_tot_unemp <- nrow(gss_re %>% filter(unemp == "Yes")) ## total number of unemployed

## calculate expected amount of unemployed per etnic group and store in
## etnic_groups dataset

etnic_groups %>%
  mutate(expected_counts = p * n_tot_unemp) -> expected_df

## calculate observed amount of unemployed per etnic group and store in
## observed_df dataset

gss_re %>% 
  filter(unemp == "Yes") %>%
  group_by(race) %>%
  summarise(observed_p = n()) -> observed_df

## join the two dataset in a single one

final_df <- full_join(observed_df, expected_df, by="race")

final_df
```

### 5.2 Inference

Hypothesis

**Null hypothesis**: The distribution of unemployement follows the hypothesized 
distribution. The hypothesized distribution is the overall distribution of 
people belonging to different etnic groups.

**Alternate hypothesis**: The distribution of observed proportion of unemployed
does not follow the hypothesized distribution.

we will test theese hypothesis with the chi-square goodness of fit test. First
we have to check if the conditions for applying the test are satisfied.

Check Conditions

1. **Independence**: the GSS dataset is collected from a random sample survey. 
We safely assume that the interviews are independent.

2. **Expected counts**: the expected count for each cell should be higher than
a minimum of 5. This condition is met.

3. **Degrees of Freedom**: the degrees of freedom should be at least 2. In our
study the degrees of freedom is 3-1 = 2.

Conditions are met, we can proceed with the chi-squared test.

### 5.3 Chi-Squared goodness of fit

```{r}
chisq.test(x = final_df$observed_p, p = final_df$p)```
```

The chi-squared value is 168.36 with a p-value of almost zero. We can reject the
null hypothesis, the observed distribution of unemployement rates among different
etnic groups doesn't follow the expected distribution.

### 5.4 Finding

We have found evidence that the unemployment rate between different etnic groups
is not equally distribuited, with more observed than expected proportion of
unemployed between other than "White" etnic group.

* * * 

## 6. Income disparity baser on region of interview

Variables analyzed to answer this question:

1. **region**: A categorical variable, a nine-point scale, that classifies the
region of interview into 1 - New England, 2 - Middle Atlantic, 3 - East North
Central, 4 - West North Central, 5 - South Atlantic, 6 - East South Central,
7 - West South Central, 8 - Mountain, 9 - Pacific

2. **coninc**: Inflation-adjusted total family income.

We prepare first a simpler dataset, and remove null values:

```{r}
gss %>%
  filter(!is.na(region) & !is.na(coninc)) %>%
  select(region, coninc)  -> gss_reg

dim(gss_reg)
```

We have now 51232 rows from the starting 57061.
 
### 6.1 Exploratory Data Analysis

```{r}
table(gss_reg$region)
```

Quickly checking the above table, we see that we have a fair amount of observation
in each of the 9 regions. Let's see how the total income distribution looks like
for each of the 9 regions:

```{r}
ggplot(data = gss_reg, aes(x = coninc)) + geom_histogram() + facet_wrap(~region)
```

**Observations:**

1. The shape of the distribution of income seems to be the same, with a strong
right schewness. Even if the size of the sample is different from region to
region, with a minimum in New England and a maximum in South Atlantic,
(see file Income_per_Regions.pdf).

### 6.2 Inference

Hypothesis

**Null hypothesis:** The mean of total family income with 
inflation correction is the same across different regions.

**Alternative hypothesis:** The mean of total family income with 
inflation correction is the different between at least two regions.

We have to compare the mean value between more than 2 groups of citizenship, 
so we will use the ANOVA analysis. First we check the necessary conditions in
order to perform the ANOVA analysis.

Check Conditions

1. **Independence**: the GSS dataset is collected from a random sample survey. 
We safely assume that the interviews are independent.

2. **Nearly normal distributions within each group**: To verify the normality of
distribution, we explore the quantile-quantile plot for each of the nine cases.

```{r}
# 9 graphs in 3 rows
par(mfrow = c(3,3))

# Iterate over the groups and graph a QQ plot to test normality
groups = c("New England", "Middle Atlantic", "E. Nor. Central",
	"W. Nor. Central", "South Atlantic", "E. Sou. Central", 
	"W. Sou. Central", "Mountain", "Pacific")

for (i in 1:9) {
	df = gss_reg %>% filter(region == groups[i])  
	qqnorm(df$coninc, main = groups[i])
	qqline(df$coninc)
}
```
*Observations:*

There is a significant deviation from standard normal distribution in each of the
nine groups especially in the upper quantile, which is consistent with the
strong skewness of the distributions, (see the file
Regions_and_Income_normality_check.pdf)

3. **Variability**: The variability across the groups needs to be about equal. 

```{r}
ggplot(data = gss_reg, aes(x = region, y = coninc)) + 
  geom_boxplot(aes(fill = region))
```

The variability seems consistent between the nine groups.

Based on the above observations we can perform the ANOVA analysis, keeping in mind 
condition 2 was not satisfied, so we need to be cautious in interpreting 
the results of ANOVA.

### 6.3 ANOVA

```{r}
anova(lm(coninc ~ region, data = gss_reg))
```

We have a F-statistics of 90.335 and p-values almost zero. We have strong evidence
that at least a pair of region groups have different main total inflation corrected
family income, that cannot be explained simply by sampling variability.
To analize which pair of groups has different income we Type I error.

R has the following function tha performs pairwise t-tests with the 
Bonferroni correction.

```{r}
pairwise.t.test(gss_reg$coninc, gss_reg$region, p.adj="bonferroni")
```

*Observations:*

Many pairs of values have a p-value close to 1. So we have evidence that the mean
total inflation corrected family income is different in different regions of 
the U.S. and this difference cannot be explained by the sampling variability. 

### Findings

Although ANOVA conditions were not fulfilled, as we observed from exploratory data 
analysis, we observed the income to be different between regions.
Strongest difference is between following pairs:

"Pacific" and "Middle Atlantic"
"Mountain" and "E. Nor. Central"
"South Atlantic" and "W. Nor. Central"
"W. Sou. Central" and "W. Nor. Central"
"Mountain" and "W. Nor. Central"
"Mountain" and "New England"

* * * 

## Conclusion

In this project, it was used the Social Survey Coursera Extract dataset which
was a reduced version of the original, offered by the Coursera course in
Inferential Statistics by Duke University.
I drew statistical inferences for each research question and the resultscan 
be generalized to the entire U.S. population.


